---
title: "Benchmarking Categorical Clustering Methods Across Diverse Datasets"
author: "Mahmoud Alwaraqi, Senior Data Scientist"
date: "`r format(Sys.time(),'%B %d, %Y')`"
output:
  pdf_document:
    citation_package: natbib
    keep_tex: no
    fig_caption: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
header-includes:
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
always_allow_html: TRUE
bibliography: master.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

Categorical data appears across a wide range of diverse industries. In market research and e-commerce, examples include  survey data, point-of-sale (POS) data, longitudinal customer transaction data, and subscription preferences. In finance and banking, examples include credit card applications, insurance claim logs, and investment profiles based on risk tolerance, financial goals, etc.  Categorical data is also highly relevant in the healthcare and medicine industries where data consists of symptom checklists, patient intake records, and clinical trial profiles. All of these industries, and more, benefit from utilizing clustering methods. The data may consist of nominal (e.g., color) or ordinal variables (e.g., income brackets) and exclude continuous variables from the analysis to focus the scope of this paper on categorical data. The framework of some of the methods mentioned here can easily extend to continuous variables, even without recoding them, such as Finite Mixutre Models (FMMs).

In this paper I compare clustering methods on different categorical datasets. The datasets I experiment with in this paper are the following:

* Iris
* Mushroom
* Zoo

The methods I use for clustering categorical data include the following: K-Modes, Hierarchical clustering, and Finite Mixture Models (FMMs) including Latent Class Models (LCMs).  

This paper demonstrates the advantages of Finite Mixture Models as they tend to give consistent quality performance and outperform other methods. 

# Overview of the methods

## K-Modes

k-modes clustering is an extension of k-means clustering, where the mode
is used to define the center of the cluster instead of the mean. Recall
that the mode is defined as the most frequently occurring value. It's
useful when working with categorical data but can be relatively more
computationally intensive, especially as the number of variables
increases.

The steps are the following [@MalikTuckfield2019AUL]

1.  Choose any $k$ number of random points as cluster centers.

2.  Find the Hamming distance of each point from the center.

3.  Assign each point to a cluster whose center it is closest to.

4.  Choose new cluster centers in each cluster by finding the mode of
    all data points in that cluster.

5.  Repeat from step 2 until the cluster centers stop changing.

To pick the number of clusters we will use the average silhouette
scores/widths. For each data point $i$ with features $y_i \in C_I$ the
silhouette is

$$
s(i)=\frac{b(i)-a(i)}{\max \{a(i), b(i)\}} \text {, if }\left|C_I\right|>1
$$ where $$
a(i)=\frac{1}{\left|C_I\right|-1} \sum_{y_k \in C_I, k \neq i} d(y_i, y_k)
$$

$$
b(i)=\min _{J \neq I} \frac{1}{\left|C_J\right|} \sum_{y_j \in C_J} d(y_i, y_j)
$$ $a(i)$ is the average distance of point $i$ to other points in its
cluster. $b(i)$ is the minimum of the set of average distances between
point $i$ and other other points in different clusters. $s(i)$ ranges
between $-1$ and $1$, higher values mean the observation is well matched
to its cluster. Given any number of clusters, $k$, the average of $s(i)$
across all observations will be an indicator of whether or not $k$ is
good. A value above 0.70 is considered best, reasonable above 0.50, and
weak above 0.25. we will use the `klaR` package to run k-modes on our
data.

## Hierarchical clustering (HC)
Hierarchical clustering is a tree-based method. There are two flavors of HC:

1. Agglomerative, which starts with all data points assigned their own clusters, and 

2. Divisive, which starts with all data points assigned to a single cluster

HC is similar to k-modes in that we must decide on the number of clusters based on silhouettes, the elbow method, or visual inspection of the dendrogram.  However, since we're using using categorical data we need to decide on a measure of distance which we take to be Gower's distance with weights set to 1, which is simply the total number of mismatches.

$$d(i,j) = \frac{\sum_{k=1}^{n} w_k \cdot \delta_{ijk} \cdot d_{ijk}}{\sum_{k=1}^{n} w_k}$$

Where:

- **$n$** is the total number of features/variables in the dataset.
- **$w_k$** is the weight assigned to feature $k$ (typically set to $1$ by default).
- **$\delta_{ijk}$** is a binary indicator ($0$ or $1$) that drops to $0$ if either observation has a missing value (`NA`) for feature $k$, effectively ignoring it.
- **$d_{ijk}$** is the specific partial distance calculated for feature $k$ based on its specific data type.

However, as we move into higher dimensional data, Gower's distance fails and it's likely we'd need to resort to Multiple Correspondence Analysis (MCA) before running HC, this method is known as HCPC, or hierarchical clustering on principal components.

## Finite Mixture Models (FMMs) 
Probability-based cluster modeling has several advantages, as argued by
@BouveyronCeleuxMurphyRaftery2019MCCDS. Although I will not list those
advantages here, I'll say that probability-based clustering seems to
perform comparably well to, and in many cases better than, non-probability models. Broadly, the FMMs is the following:

$$f(x) = \sum_{k=1}^{K} \pi_k f_k(x)$$
where 

* $\pi_k$ represents the **mixing proportions** (the probability that a 
randomly selected observation belongs to cluster $k$).
* $f_k(x)$ represents the **component density** (the probability distribution 
specific to cluster $k$).

As with other clustering methods, in an unsupervised learning context, the total number of clusters, $K$, will need to be estimated.

### All nominal variables
When the categorical variables are all nominal, we use Latent Class Models (LCMs) as formulated in the `Rmixmod` R package.  Latent class analysis was first introduced by the Mathematical Sociologist Paul F. Lazarsfeld to analyze survey response patterns. 

The set of responses from observation $i$ with $d$ categorical variables
is denoted by $y_i$, each variable with, potentially, a different number
of levels. $y_i$ is coded by the binary vector
$\left( y_{i}^{jh};j=1,...,d;h=1,...,m_j\right)$ with

$$
y_i^{j h}=\begin{cases}1 & \text { if } y_i^j=h \\ 0 & \text { otherwise. }\end{cases}
$$

Data are modeled as a mixture of *G* multivariate multinomial
distributions with probability distribution function
[@BouveyronCeleuxMurphyRaftery2019MCCDS]

$$
f\left(y_i ; \theta\right)= \sum_{g=1}^G \tau_g \prod_{j, h}\left(\alpha_g^{j h}\right)^{y_i^{j h}}
$$ where $\alpha_g^{jh}$ is the probability that variable $j$ has level
$h$ if observation $i$ is in cluster $g$. And $\tau_g$ are the mixing
proportions of the $G$ latent clusters.

The log-likelihood is

$$
\mathcal{L}(\theta)=\mathcal{L}(\theta ; y)=\sum_{i=1}^n \log \left(\sum_{g=1}^G \tau_g \prod_{j=1}^d \prod_{h=1}^{m_j}\left(\alpha_g^{j h}\right)^{y_i^{j h}}\right)
$$

Conditioning on the latent *class* (i.e., cluster), $z = (z_1,...,z_G)$
with $z_g = (z_{1g},...,z_{ng})$ and $z_{ig} = 1$ if $y_i$ is from
cluster $g$ and $z_{ig} = 0$ otherwise, we get the complete-data
log-likelihood

$$
\mathcal{L}_C(\theta)=\mathcal{L}(\theta ; y, z)=\sum_{i=1}^n \sum_{g=1}^G z_{i g} \log \left(\tau_g \prod_{j=1}^d \prod_{h=1}^{m_j}\left(\alpha_g^{j h}\right)^{y_i^{j h}}\right)
$$ The complete-data log-likelihood is then reparameterized to define a
family of five parsimonious LCMs based on five constraints on the
parameters, as the following[^1]:

[^1]: Technically, by "$\alpha$-probability" in the list of these five
    models I mean specifically the "scatter" parameter that results from
    the reparameterization of the complete-data log-likelihood which is
    not shown in this article. Essentially, the "scatter" is a
    reparameterizaiton of the probability $\alpha_g^{jh}$ that reflects
    how the probabilities of the levels of a variable could differ from
    the probability of the mode value of said variable. If you'd like
    more details please see @BouveyronCeleuxMurphyRaftery2019MCCDS.

1.  The $\alpha$-probability depends upon clusters, variables, and
    variable levels.

2.  The $\alpha$-probability depends upon clusters, variables, and
    **not** variable levels.

3.  The $\alpha$-probability depends upon clusters, **not** variables.

4.  The $\alpha$-probability depends upon variables, **not** clusters,
    and **not** variable levels.

5.  The $\alpha$-probability is constant over variables and clusters.

Additionally, the mixing proportions, $\tau_g$, can either be free or
constant, with $\tau_g = p$ for all $g$. This leads to ten different
constraints on the parameters. The ICL criterion will be used to select
the best model and the number of clusters for it. @BouveyronCeleuxMurphyRaftery2019MCCDS favors
ICL for clustering and BIC for density estimation.

### Nominal and/or all ordinal
We use a latent threshold model formulation as implemented in the the `clustMD` R package.

The package handles ordinal variables by projecting them into an unobserved, continuous **latent space**. Let $\mathbf{z}_i = (z_{i1}, \dots, z_{iJ})'$ represent the hidden continuous vector for observation $i$ across $J$ dimensions. If observation $i$ is assigned to latent cluster $g$, its distribution is modeled as a multivariate Gaussian density:

$$\mathbf{z}_i \mid \text{Cluster } g \sim \text{MVN}(\boldsymbol{\mu}_g, \boldsymbol{\Sigma}_g)$$

To maintain computational feasibility during Monte Carlo EM optimization, the package restricts the latent covariance matrix $\boldsymbol{\Sigma}_g$ to a diagonal structure:

$$\boldsymbol{\Sigma}_g = \text{diag}(\sigma_{g1}^2, \sigma_{g2}^2, \dots, \sigma_{gJ}^2)$$

This means that within any single cluster, the latent variables are assumed to be independent, while their overall geometry is modified across groups according to the chosen `model` argument:

*   **`model = "EII"` (Spherical, Equal Volume):** Constrains the covariance to:
    $$\boldsymbol{\Sigma}_g = \sigma^2 \mathbf{I}_J$$
    All variables share an identical variance, forcing clusters into rigid, circular shapes of equal volume across all groups.
    
*   **`model = "VII"` (Spherical, Variable Volume):** Constrains the covariance to:
    $$\boldsymbol{\Sigma}_g = \sigma_g^2 \mathbf{I}_J$$
    Variables within a cluster share the same variance, but the total volume (the scale of $\sigma_g^2$) can vary freely between different clusters.
    
*   **`model = "EEI"` (Diagonal, Equal Volume & Shape):** Constrains the covariance to:
    $$\boldsymbol{\Sigma}_g = \boldsymbol{\Sigma} = \text{diag}(\sigma_1^2, \sigma_2^2, \dots, \sigma_J^2)$$
    Variables can exhibit unique variances from one another (producing elliptical clusters), but this exact shape and volume profile is held constant across all groups.
    
*   **`model = "VVI"` (Diagonal, Variable Volume & Shape):** Unrestricted diagonal covariance given by:
    $$\boldsymbol{\Sigma}_g = \text{diag}(\sigma_{g1}^2, \sigma_{g2}^2, \dots, \sigma_{gJ}^2)$$
    This is the most flexible model, allowing every single variable in every individual cluster to scale its variance independently, yielding unique elliptical shapes and sizes for each group.


# Measuring performance
I use the **homogeneity score** to measure model performance. It is an information-theoretic cluster validation metric used to evaluate unsupervised machine learning models; specifically for this paper, predicted clusters against the true classes.
A model satisfies perfect homogeneity ($h = 1.0$) if all of its predicted clusters contain only data points that are members of a single class. 

The metric ignores over-clustering; it does not penalize a model for splitting a true class into multiple, highly pure sub-clusters. For example, our true class may be binary: customer purchased or not. However, the clustering model identifies  patterns such as cluster 1 = "18 to 24 year old's who browsed the online gaming section for more than 10 minutes" as a cluster.  Say there are 15 such patterns/subgroups that the clustering model identities.  We are concerned with the purity/homogeneity of these clusters; that is, being entirely, or mostly, "purchased" or "not purchased". If individuals with attributes in cluster 1 is close to 100% or 0% purchased then that is a good thing; what we don't want is something like 50% purchased in cluster 1. We are **not** concerned with, and will not penalize, the clustering model for finding 15 of these patterns/subgroups instead of only the 2, purchased vs not-purchased. What we care about is how well each cluster identifies with a single class.




## Mathematical Formulation

The formulation relies entirely on Shannon Entropy and Conditional Entropy to measure how much uncertainty about the true class labels ($C$) is removed given the predicted cluster assignments ($K$).

### The Core Formula

The formal mathematical definition of the homogeneity score ($h$) is:

$$h = 1 - \frac{H(C \mid K)}{H(C)}$$

Where:
* $H(C)$ is the baseline entropy of the ground-truth classes.
* $H(C \mid K)$ is the conditional entropy of the classes given the predicted clusters.

### True Class Baseline Entropy $H(C)$

This measures the inherent uncertainty or structural diversity of the true classes across the entire dataset:

$$H(C) = - \sum_{c=1}^{|C|} \frac{n_c}{n} \log\left(\frac{n_c}{n}\right)$$

where:  

* $n$: Total number of data points in the dataset.  
* $n_c$: Number of data points belonging to true class $c$.   
* $\frac{n_c}{n}$: The probability that a randomly selected point belongs to class $c$.   


### Conditional Entropy $H(C \mid K)$

This tracks the remaining uncertainty about the true classes *after* inspecting the contents of the predicted clusters:

$$H(C \mid K) = - \sum_{k=1}^{|K|} \sum_{c=1}^{|C|} \frac{n_{c,k}}{n} \log\left(\frac{n_{c,k}}{n_k}\right)$$

Where:

* $n_k$: Total number of points assigned to predicted cluster $k$.  
* $n_{c,k}$: Number of points that belong to true class $c$ and were assigned to cluster $k$.  
* $\frac{n_{c,k}}{n_k}$: The proportion of cluster $k$ that belongs to class $c$ (representing cluster purity).  


## Boundary Behaviors

* **Perfect Purity ($h = 1.0$):** If every cluster contains points from only one true class, then $\frac{n_{c,k}}{n_k} = 1$ for that class and $0$ for all others. Because $\log(1) = 0$, the conditional entropy drops to zero ($H(C \mid K) = 0$), yielding $h = 1 - 0 = 1.0$.

* **Random Clustering ($h \approx 0.0$):** If a cluster assignment provides zero structural information, the distribution of classes inside each cluster mirrors the overall dataset. Thus, $H(C \mid K) = H(C)$, yielding $h = 1 - 1 = 0.0$.


```{r faster_render_params, include=FALSE}
# I set certain internal parameters that help me to render the R markdown faster by,
# for example, skipping certain steps that I already ran. 
# This is useful so that I don't need to rerun laborious steps when rendering the document.

run_iris_fmm_modelselect <- FALSE 
run_iris_fmm_best_model <- FALSE

run_shroom_kmodes_bagging <- FALSE

run_shroom_lcm_models <- FALSE
```

```{r libraries, warning=FALSE, include=FALSE}
# library(magrittr)
library(knitr)
library(MASS)
library(klaR)
library(ggplot2)
library(gridExtra)
library(FactoMineR)
library(factoextra)
library(corrplot)
library(cluster)
library(ClusterR) # For "mushroom" data
library(GDAtools)
library(Rmixmod)
library(caret)
library(kableExtra)
library(mclust)
library(clue)
library(clustMD)
library(future)
library(furrr)
library(vcd)
library(viridis)
main_dir <- dirname(rstudioapi::getSourceEditorContext()$path)
intermediates_dir <- file.path(main_dir,"data")
if (!dir.exists(intermediates_dir)){
  dir.create(intermediates_dir,recursive = TRUE)
}
```

```{r helper_functions, message=FALSE, warning=FALSE, include=FALSE}
plot_cluster_categorical_flexible <- function(
    data, cluster_col, cat_col, use_gradient = FALSE,title = NA_character_) {
  
  plot_data <- data |>
    dplyr::count(.data[[cluster_col]], .data[[cat_col]]) |>
    dplyr::group_by(.data[[cluster_col]]) |>
    dplyr::mutate(percentage = n / sum(n) * 100)
  
  plot_title <- if (!is.na(title)) title else paste("Distribution of", cat_col, "by Cluster")
  p <- ggplot(plot_data, aes(x = .data[[cluster_col]], y = percentage, fill = .data[[cat_col]])) +
    geom_bar(stat = "identity", position = "stack", color = "white", linewidth = 0.3) + 
    theme_minimal() +
    labs(
      title = plot_title,
      x = "Clusters", 
      y = "Percentage (%)"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x = element_blank()
    )
  
  # Toggle color behavior based on argument
  if (use_gradient) {
    num_levels <- length(unique(plot_data[[cat_col]]))
    level_colors <- colorRampPalette(c("#DCEEFF", "#08306B"))(num_levels)
    p <- p + 
      scale_fill_manual(values = level_colors) + 
      labs(fill = "Levels (Light to Dark)")
  } else {
    p <- p + 
      scale_fill_viridis_d(option = "turbo") + 
      labs(fill = "Categories")
  }
  
  return(p)
}
cluster_categorical_heatmap <- function(data,cluster_col,cat_cols,title = NA_character_){
  long_df <- data |>
    dplyr::select(dplyr::all_of(c(cluster_col,cat_cols))) |> 
    tidyr::pivot_longer(
      cols = cat_cols,
      names_to = "Variable",
      values_to = "Category_Level"
    ) |>
    dplyr::count(!!rlang::sym(cluster_col), Variable, Category_Level) |>
    dplyr::group_by(!!rlang::sym(cluster_col), Variable) |>
    dplyr::mutate(Proportion = n / sum(n)) |> 
    # Convert Variable to a factor with levels matching cat_cols order
    dplyr::mutate(Variable = factor(Variable, levels = cat_cols))
  
  plot_title <- if (!is.na(title)) title else "Cluster Profiles Across Categorical Variables"
  
  # Plot a grid of heatmaps 
  ggplot(long_df, aes(x = !!rlang::sym(cluster_col), y = Category_Level, fill = Proportion)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlGnBu", direction = 1) +
    facet_wrap(~Variable, scales = "free_y") + # Handles varying levels (2 to 10) per plot
    theme_minimal() +
    labs(
      title = plot_title,
      x = "Clusters",
      y = "Category Levels"
    ) +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      strip.background = element_rect(fill = "gray90")
    )  
}
predict_kmodes <- function(model, newdata) {
  # Custom predict function for k-modes (finds closest mode via simple matching distance)
  modes <- model$modes  # Matrix of cluster modes
  num_clusters <- nrow(modes)
  num_obs <- nrow(newdata)
  
  # Initialize matrix to store distance from each row to each cluster mode
  distances <- matrix(0, nrow = num_obs, ncol = num_clusters)
  
  # Calculate simple matching distance (count mismatches)
  for (k in 1:num_clusters) {
    # Compare each row of newdata to the k-th cluster mode profile
    mismatches <- t(apply(newdata, 1, function(row) row != modes[k, ]))
    distances[, k] <- rowSums(mismatches)
  }
  
  # Assign to the cluster with the minimum distance (fewest mismatches)
  assigned_clusters <- apply(distances, 1, which.min)
  return(assigned_clusters)
}
calc_cramers_matrix <- function(df) {
  n <- ncol(df)
  mat <- matrix(1, nrow = n, ncol = n, dimnames = list(names(df), names(df)))
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      stat <- vcd::assocstats(table(df[[i]], df[[j]]))
      mat[i, j] <- stat$cramer
      mat[j, i] <- stat$cramer
    }
  }
  return(mat)
}

```

# Iris dataset

The Iris dataset is a classic data frame introduced by British statistician Ronald Fisher in 1936, it is often used sanity-check basic functionality. 
It consists of 150 observations split evenly across three distinct species of Iris flowers, Setosa, Versicolor, and Virginica. It comprises samples of four measurements:  Sepal.Length, Sepal.Width, Petal.Length, and Petal.Width.  For testing, we can pretend we don't know the species and use the four measurements to guess it. Since we're interested in clustering with categorical variables, let's recode the four numeric measurement variables. 
variables.


```{r iris_percentiles, echo=FALSE}
iris_measurement_variables <- base::setdiff(colnames(iris),"Species")
names(iris_measurement_variables) <- iris_measurement_variables

iris_quantiles <-
purrr::map(iris_measurement_variables,
           function(x){
             iris |> 
               dplyr::summarise(tibble::as_tibble_row(quantile(!!rlang::sym(x),probs=c(0.05,0.25,0.50,0.75,0.95))))
           }
           )

iris_quantiles_df <- dplyr::bind_rows(iris_quantiles) 
iris_quantiles_df$variable <- names(iris_quantiles)
iris_quantiles_df <- dplyr::relocate(iris_quantiles_df,variable)
knitr::kable(iris_quantiles_df,digits=1,col.names = c("",colnames(iris_quantiles_df)[-1]),caption="Iris percentiles")
```

We will create four levels for each variable based on the values of the quartiles. 

```{r echo=FALSE,warning=FALSE}
iris_no_species <-
  iris |> 
  dplyr::select(-Species) |> 
  dplyr::mutate(sepal_length = dplyr::case_when(Sepal.Length < 5.1 ~ "Low",
                                                Sepal.Length >= 5.1 & Sepal.Length < 5.8 ~ "Medium",
                                                Sepal.Length >= 5.8 & Sepal.Length < 6.4 ~ "Medium-High",
                                                Sepal.Length >= 6.4 ~ "High"
                                                  ),
                
                sepal_width = dplyr::case_when(Sepal.Width < 2.8 ~ "Low",
                                               Sepal.Width >= 2.8 & Sepal.Width < 3.0 ~ "Medium",
                                               Sepal.Width >= 3.0 & Sepal.Width < 3.3 ~ "Medium-High",
                                               Sepal.Width >= 3.3 ~ "High"
                                                      ),
                
                petal_length = dplyr::case_when(Petal.Length < 1.60 ~ "Low",
                                                Petal.Length >= 1.60 & Petal.Length < 4.35 ~ "Medium",
                                                Petal.Length >= 4.35 & Petal.Length < 5.10 ~ "Medium-High",
                                                Petal.Length >= 5.10 ~ "High"),
                
                petal_width = dplyr::case_when(Petal.Width < 0.3 ~ "Low",
                                               Petal.Width >= 0.30 & Petal.Width < 1.30 ~ "Medium",
                                               Petal.Width >= 1.30 & Petal.Width < 1.80 ~ "Medium-High",
                                               Petal.Width >= 1.80 ~ "High")
                )
                
iris_recode <-
  iris_no_species |> 
  dplyr::select(sepal_length, sepal_width, petal_length, petal_width) |> 
  dplyr::mutate(dplyr::across(dplyr::everything(), \(x) factor(x, levels = c("Low", "Medium", "Medium-High", "High"))))

```

The data is recoded to "Low","Medium","Medium-High","High".

```{r echo=FALSE, warning=FALSE}
iris_recode_col_names <-  colnames(iris_recode) 
names(iris_recode_col_names) <- iris_recode_col_names

iris_counts_of_recode <-
  purrr::map(iris_recode_col_names,
            function(x){
              iris_recode |> 
                dplyr::count(!!rlang::sym(x)) |> 
                dplyr::arrange(!!rlang::sym(x)) |> 
                tidyr::pivot_wider(names_from = x,
                                   values_from = "n") |> 
                dplyr::mutate(variable = x)
              
            })

iris_counts_of_recode_df <- 
  dplyr::bind_rows(iris_counts_of_recode) |> 
  dplyr::relocate(variable)

knitr::kable(
  iris_counts_of_recode_df,
  col.names = c("Recode Variable",colnames(iris_counts_of_recode_df)[-1]),
  caption = "Counts of four iris categorical variables"
  )

```

## Iris: k-modes

We use `factoextra::fviz_nbclust()` to get silhouette scores for 8
clusters. This function returns a plot of average silhouette scores by cluster, among other things.

If we ignore the ordinal nature of the measurement variables, which we can do by specifying the Gower distance in the `diss` argument of `factoextra::fviz_nbclust()`, then k-modes fails badly.

```{r echo=TRUE, warning=FALSE}
gower_dist <- cluster::daisy(iris_recode, metric = "gower")
set.seed(45)
iris_k_modes_silhouette_nom <- factoextra::fviz_nbclust(
  iris_recode,
  FUNcluster = klaR::kmodes,
  method = "silhouette",
  diss = gower_dist,
  k.max = 8
  )

iris_recode_integers <- iris_recode |> 
    dplyr::mutate(dplyr::across(dplyr::everything(), \(x) as.integer(x)))
set.seed(45)
iris_k_modes_silhouette_ord <- factoextra::fviz_nbclust(
  iris_recode_integers,
  FUNcluster = klaR::kmodes,
  method = "silhouette",
  k.max = 8
  )
```

```{r echo=FALSE}
grid.arrange(
  iris_k_modes_silhouette_nom,
  iris_k_modes_silhouette_ord,
  ncol = 2,
  top = "Iris recode Silhouette scores: Gower (Left) vs Euclidean (Right)"
  )
```

Treating the measurements naturally, as ordinal variables, suggests 3 clusters, while ignoring the ordinal nature of the measurements suggests more than twice the number of clusters.

Know that passing the data as integers or as factors into `klaR::kmodes()` doesn't really matter in general, and certainly not for us here, since the function uses the Hamming distance which is based on matches and mismatches of values.

```{r echo=TRUE}
## Optimal number of clusters if treating as nominal
set.seed(45)
iriscat_kmodes_nom_45 <- klaR::kmodes(
  iris_recode,7,iter.max = 1000
  )
## Optimal number of clusters if treating as ordinal
set.seed(45)
iriscat_kmodes_ord_45 <- klaR::kmodes(
  iris_recode,3,iter.max = 1000
  )
```

```{r echo=FALSE}
iris_performance <- dplyr::tibble(
  method_name = "k-modes",
  variables = "nominal",
  parameters = "k=7",
  rand.seed = "45",
  homo_score = clevr::homogeneity(iris$Species, iriscat_kmodes_nom_45$cluster) |> round(2)
  )
iris_performance <- iris_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "k-modes",
      variables = "ordinal",
      parameters = "k=3",
      rand.seed = "45",
      homo_score = clevr::homogeneity(iris$Species, iriscat_kmodes_ord_45$cluster) |> round(2)
    )
  )
knitr::kable(
  iris_performance,
  caption="Iris: Methods performance")
```

Since k-modes, like k-means, can be sensitive to the (random) starting point, we will try a
different (random) starting point, `set.seed(323)`.

```{r echo=FALSE}
set.seed(323)
iriscat_kmodes_323 <- klaR::kmodes(
  iris_recode,3,iter.max = 1000)
iris_pred_table <- table(iris$Species,iriscat_kmodes_323$cluster)
knitr::kable(
  iris_pred_table,
  caption="Iris recode: 3-cluster k-modes starting at `set.seed(323)`.")
iris_performance <- iris_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "k-modes",
      variables = "ordinal",
      parameters = "k=3",
      rand.seed = "323",
      homo_score = clevr::homogeneity(iris$Species, iriscat_kmodes_323$cluster) |> round(2)
      )
    )
set.seed(323)
iriscat_kmodes_323 <- klaR::kmodes(
  iris_recode,7,iter.max = 1000)
iris_pred_table <- table(iris$Species,iriscat_kmodes_323$cluster)
knitr::kable(
  iris_pred_table,
  caption="Iris recode: 7-cluster k-modes starting at `set.seed(323)`.")
iris_performance <- iris_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "k-modes",
      variables = "nominal",
      parameters = "k=7",
      rand.seed = "323",
      homo_score = clevr::homogeneity(iris$Species, iriscat_kmodes_323$cluster) |> round(2)
      )
    )



knitr::kable(
  iris_performance,
  caption="Iris: Methods performance")
```

This leads to an important question, what level of performance should we expect? We use bagging to provide a better estimate, using 7 cluster model that we got when ignoring the ordinal nature of the variables.

```{r iris_kmodes_bagging, echo=TRUE}
set.seed(323)
num_bootstrap <- 100
num_clusters  <- 7

boot_models <- lapply(1:num_bootstrap, function(i,x = iris_recode) {
  # Sample row indices with replacement (Bagging)
  boot_idx  <- sample(1:nrow(x), replace = TRUE)
  boot_data <- x[boot_idx, ]
  
  # Fit k-modes on bootstrap data
  km <- klaR::kmodes(boot_data, modes = num_clusters)
  
  # Predict back on the original data layout and cast to a CLUE partition
  # This standardizes the labels so they can be averaged
  
  predicted_vector <- predict_kmodes(km, x)
  
  partition_object <- as.cl_partition(predicted_vector)
  
  return(partition_object)
  
})

# Bind predictions into a single Cluster Ensemble object
ensemble_bag <- clue::cl_ensemble(list = boot_models)

# Extract Consensus via soft/hard averaging (Default: Euclidean consensus)
ensemble_bag_consensus <- clue::cl_consensus(ensemble_bag)

# Extract final hard cluster assignments
iris_kmodes_bagged_preds <- clue::cl_class_ids(ensemble_bag_consensus)

```

```{r echo=FALSE}
table(iris$Species, iris_kmodes_bagged_preds) |> 
  knitr::kable(
    caption="Bagged 7 cluster 7 modes: truth vs predicted"
  )

iris_performance <- iris_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "bagged k-modes",
      variables = "nominal",
      parameters = "k=7,iter=100",
      rand.seed = "323",
      homo_score = clevr::homogeneity(iris$Species, iris_kmodes_bagged_preds) |> round(2)
      )
    )
knitr::kable(
  iris_performance,
  caption="Iris: Methods performance")
```

So starting from a "bad seed" bagging ensured better performance relative  to any single k-mode model.

## Iris: HC
Let's start with agglomerative hierarchical clustering.
```{r iris_hc_agg, echo=TRUE}
iris_gower <- cluster::daisy(iris_recode,metric = "gower")
set.seed(323)
grid.arrange(
  factoextra::fviz_nbclust(as.matrix(iris_gower), kmeans, method = "silhouette"),
  factoextra::fviz_nbclust(as.matrix(iris_gower), kmeans, method = "wss"),
  ncol = 2,
  top = "Agglomerative Hierarchical Clustering \n
  Iris recode: Gower's Distances with weights 1"
  )
set.seed(323)
iris_hc <- stats::hclust(iris_gower, method = "ward.D2")
plot(iris_hc) 
rect.hclust(iris_hc, k = 5, border = "red")
iris_hc_classes <- stats::cutree(iris_hc, k = 5)
```

```{r echo=FALSE}
iris_performance <- iris_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "HC-AGG",
      variables = "nominal",
      parameters = "k=5,ward.D2",
      rand.seed = "323",
      homo_score = clevr::homogeneity(iris$Species, iris_hc_classes) |> round(2)
      )
    )
knitr::kable(
  iris_performance,
  caption="Iris: Methods performance")
```

Now, divisive hierarchical clustering.

```{r iris_hc_div, echo=TRUE}
set.seed(323)
iris_hc_div <- cluster::diana(iris_gower)
plot(
  iris_hc_div, which.plots = 2, labels=FALSE, 
  main = "Divisive Clustering on Categorical Data")
rect.hclust(as.hclust(iris_hc_div), k = 5, border = "red")
iris_hc_div_classes <- stats::cutree(as.hclust(iris_hc_div), k = 5)
```

```{r echo=FALSE}
iris_performance <- iris_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "HC-DIV",
      variables = "nominal",
      parameters = "k=5",
      rand.seed = "323",
      homo_score = clevr::homogeneity(iris$Species, iris_hc_div_classes) |> round(2)
      )
    )
knitr::kable(
  iris_performance,
  caption="Iris: Methods performance")
```

## Iris: LCM
We start by using the ICL criterion to find the best of 10 possible
models. We loop over each model using `Rmixmod::mixmodCluster()` and 8
clusters for each.

```{r echo=TRUE}
names_possible_models <- c("Binary_p_E","Binary_p_Ej", "Binary_p_Ek",
                           "Binary_p_Ekj","Binary_p_Ekjh","Binary_pk_E",
                           "Binary_pk_Ej","Binary_pk_Ek","Binary_pk_Ekj",
                           "Binary_pk_Ekjh")

names(names_possible_models) <- names_possible_models
set.seed(323)
iris_lcm_models <-
  purrr::map(names_possible_models,function(x){
    res <- Rmixmod::mixmodCluster(
      iris_recode,
      nbCluster = 1:8,
      dataType = "qualitative",
      model = Rmixmod::mixmodMultinomialModel(listModels = x),
      criterion = c("ICL")
      )
    res
  }
  )
```

Now we get the ICL values by cluster then plot.

```{r echo=TRUE}
set.seed(323)
iris_lcm_models_lcl <-
  purrr::map(
    names_possible_models,function(model_name){
      
      this_model_results <- iris_lcm_models[[model_name]]
      
      this_model_nbcluster <- 
        purrr::map_int(1:8,function(x = .x){
          this_model_results@results[[x]]@nbCluster[1]  
        })
      
      this_model_icl <-
        purrr::map_dbl(1:8,function(x = .x){
          this_model_results@results[[x]]@criterionValue[1]
        })*-1
      
      dplyr::tibble(num_clusters = this_model_nbcluster,
                    icl = this_model_icl) |> 
        dplyr::rename(!!model_name := icl)
    }
    
    
  )
```

```{r echo=FALSE}
iris_lcm_models_lcl_df <- purrr::reduce(
  iris_lcm_models_lcl,
  dplyr::left_join,
  by="num_clusters"
  )
# par(mar=c(1,1,1,1))

model_colors <- c(
  "Binary_p_E"="black","Binary_p_Ej"="red","Binary_p_Ek"="green",
  "Binary_p_Ekj"="blue","Binary_p_Ekjh"="purple",
  
  "Binary_pk_E"="black","Binary_pk_Ej"="red","Binary_pk_Ek"="green",
  "Binary_pk_Ekj"="blue","Binary_pk_Ekjh"="purple"
  )

model_linetypes <- c(
  "Binary_p_E"="solid","Binary_p_Ej"="solid","Binary_p_Ek"="solid",
  "Binary_p_Ekj"="solid","Binary_p_Ekjh"="solid",
  
  "Binary_pk_E"="dashed","Binary_pk_Ej"="dashed",
  "Binary_pk_Ek"="dashed","Binary_pk_Ekj"="dashed","Binary_pk_Ekjh"="dashed")

plot_base <-
  ggplot(data=iris_lcm_models_lcl_df, 
         aes(x=num_clusters))

plot_all_lines <-
  purrr::reduce(names_possible_models,function(p,model_name){

  p +
    geom_line(aes(y = rlang::eval_tidy(rlang::sym(model_name)),
                  colour = model_name),
              linetype = model_linetypes[model_name]) + 
    geom_point(aes(y =  rlang::eval_tidy(rlang::sym(model_name))),
               shape = 0)
},
.init = plot_base)

plot_final <-
  plot_all_lines +
  scale_color_manual(name = "Models",
                     values = model_colors) +
  scale_x_continuous(breaks=seq(1,8,1)) +
  xlab(label = "Number of clusters") +
  ylab(label = "") +
  labs(
    title = "ICL of 10 Latent Class Models",
    subtitle = "Iris recoded categorical data"
  ) +
  theme_minimal()   

plot_final
```

The best model is `Binary_p_Ejkh` at 3 clusters. This corresponds to the
(parsimonous) LCM with a constant mixing proportion between the 3
clusters (i.e., $\frac{1}{3}$) and with $\alpha$-probabilities (i.e.,
technically the scatter) that vary with cluster, variable, and variable
level. We pull the results of the best LCM model and compare clusters
against iris Species.

```{r echo=TRUE}
iris_lcm_models_best <- iris_lcm_models[["Binary_p_Ekjh"]]@bestResult 
```

```{r echo=FALSE}
table(iris$Species,iris_lcm_models_best@partition) |> 
  knitr::kable(caption = "Iris species vs. cluster estimated from  Binary_p_Ekjh LCM")

iris_performance <- iris_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "LCM",
      variables = "nominal",
      parameters = paste0(
        "k=",
        iris_lcm_models_best@nbCluster,
        ",",
        iris_lcm_models_best@model,collapse = ""),
      rand.seed = "323",
      homo_score = clevr::homogeneity(iris$Species, iris_lcm_models_best@partition) |> round(2)
      )
    ) |> 
  dplyr::arrange(dplyr::desc(homo_score))
knitr::kable(
  iris_performance,
  caption="Iris: Methods performance")

```

## Iris: FMM
Now, let's respect the ordinal nature of the measurement variables. The `clustMD` package uses a latent threshold (or latent continuous) framework where observed categorical variables (e.g., ordinal) are assumed to be manifestations of an underlying continuous multivariate Gaussian latent vector. Let's start with model selection.

```{r include=FALSE}
############
## Because EM algorithm has random starting point, you are likely to get stuck in a local optimal
## To get a better sense of the global optimal, repeat at different random starting points to find the highest
## log likelihood  
############
# Stabilized Single-Model Wrapper Function
fit_stable_clustMD <- function(X, G, model_type, n_starts = 10) {
  best_fit <- NULL
  best_loglik <- -Inf
  
  for (s in 1:n_starts) {
    # Set a local seed that advances sequentially per start loop
    set.seed(1000 * G + s) 
    
    fit <- tryCatch({
      clustMD(
        X = X, G = G, CnsIndx = 0, OrdIndx = ncol(X),
        Nnorms = 500, MaxIter = 100, model = model_type,
        startCL = "random"
      )
    }, error = function(e) NULL)
    
    # Track and keep only the iteration with the highest Log-Likelihood
    if (!is.null(fit) && !is.null(fit$likelihood)) {
      current_loglik <- tail(fit$likelihood, 1) # Get final converged log-lik
      if ( base::isTRUE(current_loglik > best_loglik) ) {
        best_loglik <- current_loglik
        best_fit <- fit
      }
    }
  }
  return(best_fit) # Returns the global maximum across the n_starts
}
```


```{r echo=TRUE, message=FALSE, warning=FALSE}
iris_md_matrix <- iris_recode |>
  dplyr::mutate(across(everything(), \(x) as.numeric(
    factor(x, 
           levels = c("Low", "Medium", "Medium-High", "High"), 
           ordered = TRUE
           )))) |>
  as.matrix()
# Takes 10 minutes or so to complete
# Can skip if results are already available/saved
if (run_iris_fmm_modelselect){
  # Since we have only nominal/ordinal variables,
  #  we don't need to try all 6 models
  # tested_models <- c("EII","VII","EEI","VVI","VEI","EVI")
  tested_models <- c("EII", "VII", "EEI", "VVI")
  tested_n_clusters <- 2:8
  
  # Create every unique combination of model type and number of clusters
  search_grid <- base::expand.grid(
    Model = tested_models, G = tested_n_clusters, stringsAsFactors = FALSE)
  
  # Run the iterative loop over the grid
  set.seed(323)
  model_selection_results <- purrr::map_dfr(1:nrow(search_grid), function(i) {
    current_model <- search_grid$Model[i]
    current_g     <- search_grid$G[i]
    
    #Safeguard the loop against estimation convergence errors using tryCatch
    fit <- fit_stable_clustMD(
      X = iris_md_matrix,
      G = current_g,
      model_type = current_model
      )
  
    # If model successfully converged, extract metrics; otherwise return NA
    if (!is.null(fit)) {
      data.frame(
        Model = current_model,
        Clusters = current_g,
        ICL = fit$ICLhat
      )
    } else {
      data.frame(Model = current_model, Clusters = current_g, ICL = NA)
    }
  })
best_icl <- model_selection_results |> dplyr::arrange(desc(ICL))  
dir.create(file.path(intermediates_dir,"iris","fmm"),recursive = TRUE)
saveRDS(best_icl,file.path(intermediates_dir,"iris","fmm","best_icl.rds"))
} else {
  best_icl <- readRDS(file.path(intermediates_dir,"iris","fmm","best_icl.rds"))
}
best_icl <- best_icl |>
  dplyr::filter(!is.na(ICL)) |> 
  dplyr::arrange(desc(ICL)) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
best_icl_point <- best_icl |> dplyr::slice(1)

# Construct the optimization profile plot
icl_grid_plot <- ggplot(best_icl, aes(x = Clusters, y = ICL, color = Model, group = Model)) +
  # Continuous profile lines per covariance framework
  geom_line(linewidth = 1, alpha = 0.8) +
  # Points on specific evaluation steps
  geom_point(size = 2.5) +
  # Axis configuration matching your exact cluster range
  scale_x_continuous(breaks = unique(best_icl$Clusters)) +
  # Clean, standard color palette
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Model Selection via Integrated Classification Likelihood (ICL)",
    subtitle = "Higher ICL scores indicate better group separation and parsimony",
    x = "Number of Clusters (G)",
    y = "ICL Estimate (Closer to Zero is Better)",
    color = "Covariance Structure:"
  )

# Display the clean plot
print(icl_grid_plot)
```
Some points not shown because the model didn't converge within the maximum number of iterations that we specified for the given number of clusters. The best FMM model over our grid search is `r best_icl_point$Model` with `r best_icl_point$Clusters`. 

```{r echo=FALSE, message=TRUE, warning=FALSE}
set.seed(323)
if (run_iris_fmm_best_model){
  # Fit the clustMD model with random initialization overrides
iris_fmm_best_fit <- fit_stable_clustMD(
  X = iris_md_matrix,
  G = best_icl_point$Clusters,
  model_type = best_icl_point$Model,
  n_starts = 50
  )
  saveRDS(iris_fmm_best_fit,file.path(intermediates_dir,"iris","fmm","iris_fmm_best_fit.rds"))

} else {
  
  iris_fmm_best_fit <- readRDS(file.path(intermediates_dir,"iris","fmm","iris_fmm_best_fit.rds"))
  
}

iris_performance <- iris_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "FMM",
      variables = "ordinal",
      parameters = paste0(c("k=",best_icl_point$Clusters,",",best_icl_point$Model),collapse = ""),
      rand.seed = "323",
      homo_score = clevr::homogeneity(iris$Species, iris_fmm_best_fit$cl) |> round(2)
    )
  ) |> 
  dplyr::arrange(dplyr::desc(homo_score))

iris_performance |> 
  knitr::kable(
    caption = "Iris: Model Performance"
    )

```

## Iris: Cluster charts
Although LCM chooses the correct number of clusters, the HC-DIV model results in the greatest or homogeneity, or cluster purity.
```{r echo=FALSE, message=FALSE, warning=FALSE}
iris_recode$cluster <- iris_hc_div_classes
iris_recode$true_class <- iris$Species
iris_recode <- iris_recode |> dplyr::relocate(true_class)
columns_for_clustering <- setdiff(
  colnames(iris_recode),
  "cluster")
my_plots <- purrr::map(
  columns_for_clustering,
  ~ plot_cluster_categorical_flexible(
    data = iris_recode,
    cluster_col = "cluster",
    cat_col = .x,
    use_gradient = F,
    title = .x
  ))
for( i in seq(1,length(my_plots),2) ){
  j <- min(i+1,length(my_plots))
  if (j == i) these_plots <- my_plots[c(i)] else these_plots <- my_plots[c(i,j)]
  do.call(
    "grid.arrange",
    list(
      grobs = these_plots,
      nrow = 2,
      top = "HC-DIV k=5"))
}
```

\newpage
\begin{landscape}

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width='80%', fig.align='center', fig.show='hold'}
iris_recode$cluster <- iris_hc_div_classes
iris_recode$true_class <- iris$Species
iris_recode <- iris_recode |> dplyr::relocate(true_class)
cluster_categorical_heatmap(
  data = iris_recode,
  cluster_col = "cluster",
  cat_cols = setdiff(colnames(iris_recode),"cluster"),
  title = "HC-DIV k=5"
)
iris_recode$cluster <- iris_lcm_models_best@partition
cluster_categorical_heatmap(
  data = iris_recode,
  cluster_col = "cluster",
  cat_cols = setdiff(colnames(iris_recode),"cluster"),
  title = "LCM k=3,Binary_p_Ekjh"
)
```

\end{landscape}
\newpage

# Mushroom dataset
The mushroom dataset contains physical characteristics of mushrooms (8,124 instances, 22 attributes) classified as edible or poisonous. Every single feature is categorical (e.g., cap shape, gill color), making it great for our needs.


```{r echo=FALSE}
data("mushroom",package = "ClusterR")
str_output <- capture.output(str(mushroom))
str_df <- data.frame(Structure = str_output)
knitr::kable(str_df, caption = "Structure of Mushroom Dataset")
```


```{r factor-summary-table, echo=FALSE, message=FALSE, warning=FALSE}
# 1. Exhaustive variable-specific dictionary mapping for all 22 features
mushroom_dict <- c(
  # Cap attributes
  "cap_shape_b" = "bell", "cap_shape_c" = "conical", "cap_shape_x" = "convex", 
  "cap_shape_f" = "flat", "cap_shape_k" = "knobbed", "cap_shape_s" = "sunken",
  "cap_surface_f" = "fibrous", "cap_surface_g" = "grooves", "cap_surface_y" = "scaly", "cap_surface_s" = "smooth",
  "cap_color_n" = "brown", "cap_color_b" = "buff", "cap_color_c" = "cinnamon", "cap_color_g" = "gray", 
  "cap_color_r" = "green", "cap_color_p" = "pink", "cap_color_u" = "purple", "cap_color_e" = "red", 
  "cap_color_w" = "white", "cap_color_y" = "yellow",
  
  # Structural indicators & Odor
  "bruises_t" = "bruises", "bruises_f" = "no",
  "odor_a" = "almond", "odor_l" = "anise", "odor_c" = "creosote", "odor_y" = "fishy", 
  "odor_f" = "foul", "odor_m" = "musty", "odor_n" = "none", "odor_p" = "pungent", "odor_s" = "spicy",
  
  # Gill attributes
  "gill_attachment_a" = "attached", "gill_attachment_d" = "descending", 
  "gill_attachment_f" = "free", "gill_attachment_n" = "notched",
  "gill_spacing_c" = "close", "gill_spacing_w" = "crowded", "gill_spacing_d" = "distant",
  "gill_size_b" = "broad", "gill_size_n" = "narrow",
  "gill_color_k" = "black", "gill_color_n" = "brown", "gill_color_b" = "buff", "gill_color_h" = "chocolate", 
  "gill_color_g" = "gray", "gill_color_r" = "green", "gill_color_o" = "orange", "gill_color_p" = "pink", 
  "gill_color_u" = "purple", "gill_color_e" = "red", "gill_color_w" = "white", "gill_color_y" = "yellow",
  
  # Stalk attributes
  "stalk_shape_e" = "enlarging", "stalk_shape_t" = "tapering",
  "stalk_root_b" = "bulbous", "stalk_root_c" = "club", "stalk_root_u" = "cup", 
  "stalk_root_e" = "equal", "stalk_root_z" = "rhizomorphs", "stalk_root_r" = "rooted", "stalk_root_?" = "missing",
  "stalk_surface_above_ring_f" = "fibrous", "stalk_surface_above_ring_y" = "scaly", 
  "stalk_surface_above_ring_k" = "silky", "stalk_surface_above_ring_s" = "smooth",
  "stalk_surface_below_ring_f" = "fibrous", "stalk_surface_below_ring_y" = "scaly", 
  "stalk_surface_below_ring_k" = "silky", "stalk_surface_below_ring_s" = "smooth",
  "stalk_color_above_ring_n" = "brown", "stalk_color_above_ring_b" = "buff", "stalk_color_above_ring_c" = "cinnamon", 
  "stalk_color_above_ring_g" = "gray", "stalk_color_above_ring_o" = "orange", "stalk_color_above_ring_p" = "pink", 
  "stalk_color_above_ring_e" = "red", "stalk_color_above_ring_w" = "white", "stalk_color_above_ring_y" = "yellow",
  "stalk_color_below_ring_n" = "brown", "stalk_color_below_ring_b" = "buff", "stalk_color_below_ring_c" = "cinnamon", 
  "stalk_color_below_ring_g" = "gray", "stalk_color_below_ring_o" = "orange", "stalk_color_below_ring_p" = "pink", 
  "stalk_color_below_ring_e" = "red", "stalk_color_below_ring_w" = "white", "stalk_color_below_ring_y" = "yellow",
  
  # Veil & Ring attributes
  "veil_type_p" = "partial", "veil_type_u" = "universal",
  "veil_color_n" = "brown", "veil_color_o" = "orange", "veil_color_w" = "white", "veil_color_y" = "yellow",
  "ring_number_n" = "none", "ring_number_o" = "one", "ring_number_t" = "two",
  "ring_type_c" = "cobwebby", "ring_type_e" = "evanescent", "ring_type_f" = "flaring", 
  "ring_type_l" = "large", "ring_type_n" = "none", "ring_type_p" = "pendant", 
  "ring_type_s" = "sheathing", "ring_type_z" = "zone",
  
  # Spore, Population, & Habitat attributes
  "spore_print_color_k" = "black", "spore_print_color_n" = "brown", "spore_print_color_b" = "buff", 
  "spore_print_color_h" = "chocolate", "spore_print_color_r" = "green", "spore_print_color_o" = "orange", 
  "spore_print_color_u" = "purple", "spore_print_color_w" = "white", "spore_print_color_y" = "yellow",
  "population_a" = "abundant", "population_c" = "clustered", "population_n" = "numerous", 
  "population_s" = "scattered", "population_v" = "several", "population_y" = "solitary",
  "habitat_g" = "grasses", "habitat_l" = "leaves", "habitat_m" = "meadows", 
  "habitat_p" = "paths", "habitat_u" = "urban", "habitat_w" = "waste", "habitat_d" = "woods"
)

# 2. Reshape and safely map descriptions via custom namespace prefixes
all_factor_summary <- mushroom |>
  dplyr::select(-class) |> 
  tidyr::pivot_longer(cols = everything(), names_to = "Variable", values_to = "Level") |>
  dplyr::group_by(Variable, Level) |>
  dplyr::summarise(Count = dplyr::n(), .groups = "drop_last") |>
  dplyr::mutate(Percentage = (Count / sum(Count)) * 100) |>
  dplyr::arrange(Variable, dplyr::desc(Count)) |>
  dplyr::ungroup() |>
  # Construct variable-specific lookup targets to drop collisions completely
  dplyr::mutate(
    Level = as.character(Level),
    Lookup_Key = paste0(Variable, "_", Level),
    Description = mushroom_dict[Lookup_Key],
    Level_Expanded = dplyr::if_else(!is.na(Description), paste0(Level, " (", Description, ")"), Level),
    Percentage = sprintf("%.1f%%", Percentage)
  ) |>
  dplyr::select(Variable, Level_Expanded, Count, Percentage)

# 3. Build a beautiful, nested table using kableExtra layout engines
all_factor_summary |>
  knitr::kable(
    format = "latex",
    longtable = TRUE, 
    booktabs = TRUE,
    col.names = c("Mushroom Feature", "Factor Level Value", "Count (N)", "Distribution (%)"),
    caption = "Exhaustive Summary Profile of All Categorical Features",
    align = "llrr"
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left"
  ) #|>
  # kableExtra::collapse_rows(columns = 1, valign = "top")

```

## Mushroom: Feature engineering
Unlike the Iris data, Mushrooms data is relatively more messy. This can adversely affect our clustering results by adding noise. We must clean the data before clustering.

```{r echo=FALSE}
mushroom |> dplyr::select(-class) |> 
  dplyr::summarise(dplyr::across(dplyr::everything(),\(x) sum(is.na(x)))) |> 
  tidyr::pivot_longer(dplyr::everything(),
                      names_to = "variable_name",
                      values_to = "num_missing") |> 
  dplyr::arrange(dplyr::desc(num_missing)) |> 
  knitr::kable(
    caption="Mushroom data"
  )
```

There is no missing data.

We may consider removing or regrouping variables with "low" or "high" cell counts.  Certainly, variables with constant values are of no value as they only complicate the process of find good results.

```{r echo=FALSE}
all_factor_summary$percentage <- as.numeric(gsub("%", "", all_factor_summary$Percentage)) / 100
all_factor_summary |> 
  dplyr::filter(percentage > 0.90) |> 
  knitr::kable(
    caption = "Mushroom data, more than 90% in a cell"
  )
mushroom_clean <- mushroom |> dplyr::select(-veil_type)
```

Remove **`veil_type`** since it doesn't have any variation and, therefore, doesn't add any useful information for determining clusters. Let's investigate the other variables that were flagged.

```{r echo=FALSE}
all_factor_summary |> 
  dplyr::filter(Variable == "gill_attachment") |> dplyr::select(-percentage) |> 
  knitr::kable()
all_factor_summary |> 
  dplyr::filter(Variable == "veil_color") |> dplyr::select(-percentage) |> 
    knitr::kable()
all_factor_summary |> 
  dplyr::filter(Variable == "ring_number") |> dplyr::select(-percentage) |>  
    knitr::kable()

cv_matrix <- calc_cramers_matrix(mushroom_clean |> dplyr::select(-class))
  
```


```{r correlation_plot, echo=FALSE, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
# Generate the plot optimized for R Markdown
accessible_colors <- colorRampPalette(c("#E8EFF7", "#A6C8E0", "#FE6100", "#D55E00"))(100)

corrplot(
  cv_matrix, 
  method = "color",            
  type = "upper",              # Only show top triangle to reduce visual clutter
  order = "hclust",            # Cluster similar features together automatically
  col = viridis(100),          # Color blind friendly
  # col = accessible_colors, # also color blind friendly
  is.corr = FALSE,
  cl.lim = c(0, 1),            # Cramér's V scales strictly from 0 to 1
  tl.col = "black",            
  tl.srt = 45,                
  addCoef.col = "black",       # Overlay numeric values on the plot
  number.cex = 0.65,           
  diag = FALSE
)
```

While the original Audubon Field Guide provides an unweighted checklist of traits, subsequent algorithmic analysis demonstrates that the feature space is heavily anchored by a single variable: **`odor`** [@lantz_machine_2013]. Rule-induction and decision tree models consistently leverage `odor` as the root partition due to its high mutual information content with the underlying biological structures [@lantz_machine_2013]. 

When analyzing this data in an unsupervised context (with the **`class`** column removed), the 9 unique odor labels compress into 4 primary behavior clusters based on how they naturally align with secondary physical traits:

1. **The Chemical Warnings:** Foul, fishy, spicy, and pungent mushrooms. These present high structural similarity and map almost exclusively to poisonous species.
2. **The Sweet Aromas:** Almond and anise mushrooms. These hold highly specific cap/ring structures and map uniformly to edible profiles.
3. **The Neutrals:** Odorless mushrooms. This is the largest group in the dataset. Because they lack a distinct smell, they form a massive, mixed profile that must rely on secondary features (like spore print color) to segment further.
4. **The Outliers:** Creosote and musty mushrooms. These capture tiny, niche biological variations that do not cleanly align with the other three major groups.

Therefore, initializing an unsupervised algorithm (such as K-Modes) with `modes = 4` serves logical baseline for initial feature investigation.

```{r shroom_feature_importance, echo=FALSE, message=FALSE, warning=FALSE}
# Compute Gower distance (ideal metric for purely categorical features)
gower_dist <- cluster::daisy(
  mushroom_clean |> dplyr::select(-class),
  metric = "gower")

set.seed(42)

# Cluster into a representative baseline structure (e.g., k = 4)
pam_fit <- cluster::pam(gower_dist, diss = TRUE, k = 4)
mushroom_clean$cluster_id <- as.factor(pam_fit$clustering)

# Calculate Chi-Square for every feature against the cluster labels

feature_importance <- purrr::map_dfr(
  mushroom_clean |> 
    dplyr::select(-c(class,cluster_id)) |> 
    colnames(),
  function(col) {
  test <- chisq.test(table(mushroom_clean[[col]], mushroom_clean$cluster_id))
  dplyr::tibble(
    feature = col,
    ChiSq_Score = test$statistic
  )
}) |> dplyr::arrange(dplyr::desc(ChiSq_Score))

max_corr <- purrr::map_dfr(
  mushroom_clean |> 
    dplyr::select(-c(class,cluster_id)) |>
    colnames() |> 
    sort(), 
  function(col) {
    tmp_cv_matrix <- cv_matrix[,colnames(cv_matrix) != col]
    max_val <- max(tmp_cv_matrix[col,])
    argmax <- which(tmp_cv_matrix[col,] == max_val)
    max_corr_with <- colnames(tmp_cv_matrix)[argmax]
    dplyr::tibble(
      feature = col,
      max_corr = max_val,
      max_corr_with = max_corr_with
    ) 
})

feature_importance <- feature_importance |> 
  dplyr::left_join(max_corr) 

feature_importance |> 
  dplyr::mutate(dplyr::across(where(is.numeric),\(x) round(x,digits=2))) |> 
  knitr::kable(
    caption = "Mushroom feature importance"
  )

```

The feature importance and Cramer V correlations confirms our initial suspicions based on the percentages of the variable values. We drop **`gill_attachment`**, **`veil_color`**, **`ring_number`**. Additionally, we remove **`gill_size`**, **`bruises`**. 

```{r mushroom_features, include=FALSE}
mushroom_clean <- mushroom_clean |> 
  dplyr::select(-dplyr::any_of(c(
    "gill_attachment","veil_color","ring_number","gill_size","bruises")))

mushroom_features <- mushroom_clean |> dplyr::select(-c(class,cluster_id))
colnames_mushroom_features <- colnames(mushroom_features)
true_labels_mushrooms <- mushroom_clean$class
```

```{r shroom_feature_importance_2, echo=FALSE, message=FALSE, warning=FALSE}
cv_matrix <- calc_cramers_matrix(mushroom_features)
gower_dist <- cluster::daisy(mushroom_clean, metric = "gower")
set.seed(42)
pam_fit <- cluster::pam(gower_dist, diss = TRUE, k = 4)
mushroom_features$cluster_id <- as.factor(pam_fit$clustering)
feature_importance <- purrr::map_dfr(
  colnames_mushroom_features,
  function(col) {
  test <- chisq.test(table(mushroom_features[[col]], mushroom_clean$cluster_id))
  dplyr::tibble(
    feature = col,
    ChiSq_Score = test$statistic
  )
}) |> dplyr::arrange(dplyr::desc(ChiSq_Score))

max_corr <- purrr::map_dfr(
  colnames_mushroom_features |> 
    sort(), 
  function(col) {
    tmp_cv_matrix <- cv_matrix[,colnames(cv_matrix) != col]
    max_val <- max(tmp_cv_matrix[col,])
    argmax <- which(tmp_cv_matrix[col,] == max_val)
    max_corr_with <- colnames(tmp_cv_matrix)[argmax]
    dplyr::tibble(
      feature = col,
      max_corr = max_val,
      max_corr_with = max_corr_with
    ) 
})

feature_importance <- feature_importance |> 
  dplyr::left_join(max_corr) 

feature_importance |> 
   dplyr::mutate(dplyr::across(where(is.numeric),\(x) round(x,digits=2))) |> 
  knitr::kable(
    caption = "Mushroom feature importance"
  )
```
Let's turn to model fitting. 

## Mushroom: k-modes
The Mushroom dataset is many times larger than the recoded Iris dataset we started with, with dimension `r dim(mushroom_features)`.  Increasing large data dimensions presents challenges against using k-modes which I don't think is necessary to explore, at least not in this paper. We use Huang's fast frequency initiation.

```{r echo=TRUE}
# Isolate target keys and set search grid bounds
possible_k <- 1:12
mismatch_costs <- numeric(length(possible_k))

# Vectorized loop running on fast configuration flags
for (i in seq_along(possible_k)) {
  set.seed(42)
  fit <-  klaR::kmodes(
    data = mushroom_features, 
    modes = possible_k[i], 
    fast = TRUE, # Uses Huang's fast frequency initiation
    iter.max = 20
  )
  # Sum the total matching differences across all clusters
  mismatch_costs[i] <- sum(fit$withindiff)
}

# Create the evaluation dataframe
elbow_data <- data.frame(k = possible_k, Cost = mismatch_costs)

# Instantly render the Elbow graph
ggplot(elbow_data, aes(x = k, y = Cost)) +
  geom_line(linewidth = 1, color = "grey") +
  geom_point(size = 3, color = "black") +
  theme_minimal() +
  labs(
    title = "K-Modes Scale Optimization Profile (Elbow Method)",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster Mismatch Cost"
  )
```

The elbow plot suggests 6 clusters and so we run a bagged k-modes.

```{r mushroom_kmodes_bagging, echo=FALSE, message=FALSE, warning=FALSE}
mushrooms_bagged_kmodes_dir <- file.path(intermediates_dir,"mushrooms","bagged_k-modes")
if (run_shroom_kmodes_bagging){
  set.seed(42)
  num_bootstrap <- 10
  num_clusters  <- 6
  
  boot_models <- lapply(1:num_bootstrap, function(i,x = mushroom_features) {
    # Sample row indices with replacement (Bagging)
    boot_idx  <- sample(1:nrow(x), replace = TRUE)
    boot_data <- x[boot_idx, ]
    
    # Fit k-modes on bootstrap data
    km <- klaR::kmodes(
      boot_data, fast = TRUE,
      modes = num_clusters
      )
    
    # Predict back on the original data layout and cast to a CLUE partition
    # This standardizes the labels so they can be averaged
    
    predicted_vector <- predict_kmodes(km, x)
    
    partition_object <- as.cl_partition(predicted_vector)
    
    return(partition_object)
    
  })
  dir.create(mushrooms_bagged_kmodes_dir,recursive = TRUE)
  saveRDS(
    boot_models,
    file.path(mushrooms_bagged_kmodes_dir,"boot_models.rds"))
} else {
  
  boot_models <- readRDS(
    file.path(mushrooms_bagged_kmodes_dir,"boot_models.rds")
    )
}
# Bind predictions into a single Cluster Ensemble object
ensemble_bag <- clue::cl_ensemble(list = boot_models)
# Extract Consensus via soft/hard averaging (Default: Euclidean consensus)
ensemble_bag_consensus <- clue::cl_consensus(ensemble_bag)
mushrooms_kmodes_bagged_preds <- clue::cl_class_ids(ensemble_bag_consensus)

table(true_labels_mushrooms,mushrooms_kmodes_bagged_preds) |> 
  knitr::kable(
    caption = "Mushrooms data: Bagged k=modes predicted clusters"
  )

mushroom_performance <-
dplyr::tibble(
  method_name = "bagged k-modes",
  variables = "nominal",
  parameters = glue::glue("k={num_clusters},boot={num_bootstrap}"),
  rand.seed = "42",
  homo_score = clevr::homogeneity(true_labels_mushrooms,mushrooms_kmodes_bagged_preds) |> round(2)
)
mushroom_performance |> 
  knitr::kable(
    caption = "Mushroom: Model performance"
  )

```

## Mushroom: HC
For hierarchical clustering, we first get the most representative modes then we run Gower's distance on these modes to use in the clustering algorithm. Effectively we are doing the following:

1. Get "representative" modes.
2. Perform HC on these modes, assigning each to one of $k$ clusters.
3. Map original data to one of the "representative" modes.
4. Therefore, the original data rows are mapped to one of $k$ clusters.

Let's start by finding the number of optimal clusters $k$ using 100 modes to keep the execution fast.

```{r mushroom_hc_agg, echo=FALSE, message=FALSE, warning=FALSE}
# 1. Compress factor rows down to 100 unique master categorical rows (modes)
# (Using 100 instead of 500 to keep the categorical grid execution fast)
set.seed(42)
kmodes_compress <- klaR::kmodes(
  data = mushroom_features, 
  modes = 100, 
  fast = TRUE, 
  iter.max = 20
)

# Extract the compressed text centroids
categorical_centroids <- kmodes_compress$modes

# 2. Compute the correct categorical categorical distance profile (Gower Distance)
# This handles the all-factor matching calculation safely without matrix explosion
centroid_dist <- cluster::daisy(categorical_centroids, metric = "gower")

grid.arrange(
  factoextra::fviz_nbclust(as.matrix(centroid_dist), kmeans, method = "silhouette"),
  factoextra::fviz_nbclust(as.matrix(centroid_dist), kmeans, method = "wss"),
  ncol = 2,
  top = "Agglomerative Hierarchical Clustering \n
  Mushroom: Gower's Distances with weights 1"
  )

# 3. Perform Agglomerative Hierarchical Clustering on the 100 centroids
set.seed(42)
hc_tree <- hclust(centroid_dist, method = "ward.D2")

# 4. Extract final overarching group boundaries
centroid_clusters_k5 <- cutree(hc_tree, k = 5)
centroid_clusters_k8 <- cutree(hc_tree, k = 8)

# 5. Map the final assignments cleanly back to your original uncompressed rows
mushrooms_hc_agg_k5_preds <- centroid_clusters_k5[kmodes_compress$cluster]
mushrooms_hc_agg_k8_preds <- centroid_clusters_k8[kmodes_compress$cluster]

table(true_labels_mushrooms,mushrooms_hc_agg_k5_preds) |> 
  knitr::kable(
    caption="HC-AGG,k=5,FAST"
  )
table(true_labels_mushrooms,mushrooms_hc_agg_k8_preds) |> 
  knitr::kable(
    caption="HC-AGG,k=8,FAST"
  )

mushroom_performance <- mushroom_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "HC-AGG",
      variables = "nominal",
      parameters = "k=5,FAST,100 Modes",
      rand.seed = "42",
      homo_score = clevr::homogeneity(true_labels_mushrooms,mushrooms_hc_agg_k5_preds) |> round(2)
    ),
    dplyr::tibble(
      method_name = "HC-AGG",
      variables = "nominal",
      parameters = "k=8,FAST,100 Modes",
      rand.seed = "42",
      homo_score = clevr::homogeneity(true_labels_mushrooms,mushrooms_hc_agg_k8_preds) |> round(2)
    )
  )
mushroom_performance |> 
  knitr::kable(
    caption = "Mushroom: Model performance"
  )
```


Now, divisive hierarchical clustering.

```{r mushroom_hc_div, echo=FALSE}
# 3. Perform Divisive Hierarchical Clustering on the 100 centroids
set.seed(42)
hc_tree_div <- cluster::diana(centroid_dist)

# 4. Extract final overarching group boundaries
centroid_clusters_k5_div <- cutree(hc_tree_div, k = 5)
centroid_clusters_k8_div <- cutree(hc_tree_div, k = 8)

# 5. Map the final assignments cleanly back to your original uncompressed rows
mushrooms_hc_agg_k5_preds_div <- centroid_clusters_k5_div[kmodes_compress$cluster]
mushrooms_hc_agg_k8_preds_div <- centroid_clusters_k8_div[kmodes_compress$cluster]

table(true_labels_mushrooms,mushrooms_hc_agg_k5_preds_div) |> 
  knitr::kable(
    caption="HC-DIV,k=5,FAST"
  )
table(true_labels_mushrooms,mushrooms_hc_agg_k8_preds_div) |> 
  knitr::kable(
    caption="HC-DIV,k=8,FAST"
  )

mushroom_performance <- mushroom_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "HC-DIV",
      variables = "nominal",
      parameters = "k=5,FAST,100 Modes",
      rand.seed = "42",
      homo_score = clevr::homogeneity(true_labels_mushrooms,mushrooms_hc_agg_k5_preds_div) |> round(2)
    ),
    dplyr::tibble(
      method_name = "HC-DIV",
      variables = "nominal",
      parameters = "k=8,FAST,100 Modes",
      rand.seed = "42",
      homo_score = clevr::homogeneity(true_labels_mushrooms,mushrooms_hc_agg_k8_preds_div) |> round(2)
    )
  )
mushroom_performance |> 
  knitr::kable(
    caption = "Mushroom: Model performance"
  )
```


## Mushroom: LCM

```{r mushroom_lcm_models, echo=TRUE}
mushrooms_lcm_dir <- file.path(intermediates_dir,"mushrooms","lcm")
if (run_shroom_lcm_models){
names_possible_models <- c("Binary_p_E","Binary_p_Ej", "Binary_p_Ek",
                           "Binary_p_Ekj","Binary_p_Ekjh","Binary_pk_E",
                           "Binary_pk_Ej","Binary_pk_Ek","Binary_pk_Ekj",
                           "Binary_pk_Ekjh")

names(names_possible_models) <- names_possible_models
set.seed(42)
mushroom_lcm_models <-
  purrr::map(names_possible_models,function(x){
    res <- Rmixmod::mixmodCluster(
      mushroom_features,
      nbCluster = 1:12,
      dataType = "qualitative",
      model = Rmixmod::mixmodMultinomialModel(listModels = x),
      criterion = c("ICL")
      )
    res
  })
dir.create(mushrooms_lcm_dir,recursive = TRUE)
  saveRDS(
    mushroom_lcm_models,
    file.path(mushrooms_lcm_dir,"mushroom_lcm_models.rds"))
} else {
  
  mushroom_lcm_models <- readRDS(
    file.path(mushrooms_lcm_dir,"mushroom_lcm_models.rds")
    )
}

```

Now we get the ICL values by cluster then plot.

```{r echo=FALSE}
mushroom_lcm_models_lcl <-
  purrr::map(
    names_possible_models,function(model_name){
      
      this_model_results <- mushroom_lcm_models[[model_name]]
      
      this_model_nbcluster <- 
        purrr::map_int(1:12,function(x = .x){
          this_model_results@results[[x]]@nbCluster[1]  
        })
      
      this_model_icl <-
        purrr::map_dbl(1:12,function(x = .x){
          this_model_results@results[[x]]@criterionValue[1]
        })*-1
      
      dplyr::tibble(num_clusters = this_model_nbcluster,
                    icl = this_model_icl) |> 
        dplyr::rename(!!model_name := icl)
    }
    
    
  )

mushroom_lcm_models_lcl_df <- purrr::reduce(
  mushroom_lcm_models_lcl,
  dplyr::left_join,
  by="num_clusters"
  )

model_colors <- c(
  "Binary_p_E"="black","Binary_p_Ej"="red","Binary_p_Ek"="green",
  "Binary_p_Ekj"="blue","Binary_p_Ekjh"="purple",
  
  "Binary_pk_E"="black","Binary_pk_Ej"="red","Binary_pk_Ek"="green",
  "Binary_pk_Ekj"="blue","Binary_pk_Ekjh"="purple"
  )

model_linetypes <- c(
  "Binary_p_E"="solid","Binary_p_Ej"="solid","Binary_p_Ek"="solid",
  "Binary_p_Ekj"="solid","Binary_p_Ekjh"="solid",
  
  "Binary_pk_E"="dashed","Binary_pk_Ej"="dashed",
  "Binary_pk_Ek"="dashed","Binary_pk_Ekj"="dashed","Binary_pk_Ekjh"="dashed")

plot_base <-
  ggplot(data=mushroom_lcm_models_lcl_df, 
         aes(x=num_clusters))

plot_all_lines <-
  purrr::reduce(names_possible_models,function(p,model_name){

  p +
    geom_line(aes(y = rlang::eval_tidy(rlang::sym(model_name)),
                  colour = model_name),
              linetype = model_linetypes[model_name]) + 
    geom_point(aes(y =  rlang::eval_tidy(rlang::sym(model_name))),
               shape = 0)
},
.init = plot_base)

plot_final <-
  plot_all_lines +
  scale_color_manual(name = "Models",
                     values = model_colors) +
  scale_x_continuous(breaks=seq(1,12,1)) +
  xlab(label = "Number of clusters") +
  ylab(label = "") +
  labs(
    title = "ICL of 10 Latent Class Models",
    subtitle = "Mushroom categorical data"
  ) +
  theme_minimal()   

plot_final
```

The best model is `Binary_pk_Ejkh` at 11 clusters. 

```{r echo=TRUE}
mushroom_lcm_models_best <- mushroom_lcm_models[["Binary_pk_Ekjh"]]@bestResult 
```

```{r echo=FALSE}
table(true_labels_mushrooms,mushroom_lcm_models_best@partition) |> 
  knitr::kable(caption = "Mushroom: Predicted clusters LCM Binary_pk_Ekjh")

mushroom_performance <- mushroom_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "LCM",
      variables = "nominal",
      parameters = paste0(
        "k=",
        mushroom_lcm_models_best@nbCluster,
        ",",
        mushroom_lcm_models_best@model,collapse = ""),
      rand.seed = "42",
      homo_score = clevr::homogeneity(true_labels_mushrooms, mushroom_lcm_models_best@partition) |> round(2)
      )
    ) |> 
  dplyr::arrange(dplyr::desc(homo_score))
knitr::kable(
  mushroom_performance,
  caption="Mushroom: Methods performance")

```

Of these models, LCM outperformed.

## Mushroom: Cluster charts
Clusters 4,9, and 10 have 100% purity with poisonous mushrooms and characterize the overwhelming majority of them with counts of 1296, 1728, 288, respectively.  Thinking about cluster 9, the cluster with largest number of poisonous mushrooms always had  "buff" colored gills, "evanescent" ring types, with odors being "foul", "spicy", or "fishy".  In fact, these odors are only associated with poisonous mushrooms. Additionally, "large" rings are strictly associated with poisonous mushrooms, being the only ring type of cluster 4.

```{r echo=FALSE, message=FALSE, warning=FALSE}
mushroom_features$cluster <- mushroom_lcm_models_best@partition
mushroom_features$true_class <- true_labels_mushrooms
mushroom_features <- mushroom_features |> dplyr::relocate(true_class)
mushroom_features <- mushroom_features |> dplyr::select(-cluster_id)
columns_for_clustering <- setdiff(
  colnames(mushroom_features),
  "cluster")
my_plots <- purrr::map(
  columns_for_clustering,
  ~ plot_cluster_categorical_flexible(
    data = mushroom_features,
    cluster_col = "cluster",
    cat_col = .x,
    use_gradient = F,
    title = .x
  ))
for( i in seq(1,length(my_plots),2) ){
  j <- min(i+1,length(my_plots))
  if (j == i) these_plots <- my_plots[c(i)] else these_plots <- my_plots[c(i,j)]
  do.call(
    "grid.arrange",
    list(
      grobs = these_plots,
      nrow = 2,
      top = "LCM k=11"))
  
}
```

# Zoo dataset
The Zoo dataset is a classic benchmarking dataset containing 101 animal instances evaluated across 18 total columns [@uci_zoo_1990]. It is heavily utilized for evaluating classification and unsupervised clustering tasks because its physical traits naturally group animals into distinct biological families.

```{r echo=TRUE}
zoo_data <- get(data("Zoo", package = "mlbench"))
zoo_features <- zoo_data |> 
  dplyr::select(-type) |> 
  dplyr::as_tibble() |> 
  dplyr::mutate(dplyr::across(everything(),\(x) as.factor(x))) |> 
  as.data.frame()
true_labels_zoo <- zoo_data$type
```

## Zoo: k-modes

```{r echo=FALSE}
zoo_numeric_features <- as.data.frame(lapply(zoo_features, as.numeric))
set.seed(42)
zoo_k_modes_silhouette <- factoextra::fviz_nbclust(
  zoo_numeric_features,
  FUNcluster = klaR::kmodes,
  method = "silhouette",
  k.max = 12
)
zoo_k_modes_silhouette

```
```{r echo=FALSE, message=FALSE, warning=FALSE}
set.seed(42)
num_bootstrap <- 100
num_clusters  <- 4

boot_models <- lapply(1:num_bootstrap, function(i,x = zoo_features) {
  # Sample row indices with replacement (Bagging)
  boot_idx  <- sample(1:nrow(x), replace = TRUE)
  boot_data <- x[boot_idx, ]
  
  # Fit k-modes on bootstrap data
  km <- klaR::kmodes(boot_data, modes = num_clusters)
  
  # Predict back on the original data layout and cast to a CLUE partition
  # This standardizes the labels so they can be averaged
  
  predicted_vector <- predict_kmodes(km, x)
  
  partition_object <- as.cl_partition(predicted_vector)
  
  return(partition_object)
  
})

# Bind predictions into a single Cluster Ensemble object
ensemble_bag <- clue::cl_ensemble(list = boot_models)

# Extract Consensus via soft/hard averaging (Default: Euclidean consensus)
ensemble_bag_consensus <- clue::cl_consensus(ensemble_bag)

# Extract final hard cluster assignments
zoo_kmodes_bagged_preds <- clue::cl_class_ids(ensemble_bag_consensus)


table(true_labels_zoo, zoo_kmodes_bagged_preds) |> 
  knitr::kable(
    caption="Bagged k-modes 4 clusters: truth vs predicted"
  )

zoo_performance <- dplyr::tibble(
      method_name = "bagged k-modes",
      variables = "nominal",
      parameters = "k=4,iter=100",
      rand.seed = "42",
      homo_score = clevr::homogeneity(
        true_labels_zoo, zoo_kmodes_bagged_preds) |> round(2)
      )
knitr::kable(
  zoo_performance,
  caption="Zoo: Methods performance")
```

## Zoo: HC

```{r zoo_hc_agg, secho=FALSE}
zoo_gower <- cluster::daisy(zoo_features,metric = "gower")
set.seed(42)
grid.arrange(
  factoextra::fviz_nbclust(as.matrix(zoo_gower), kmeans, method = "silhouette"),
  factoextra::fviz_nbclust(as.matrix(zoo_gower), kmeans, method = "wss"),
  ncol = 2,
  top = "Agglomerative Hierarchical Clustering \n
  Zoo: Gower's Distances with weights 1"
  )
set.seed(42)
zoo_hc <- stats::hclust(zoo_gower, method = "ward.D2")
plot(zoo_hc) 
rect.hclust(zoo_hc, k = 3, border = "red")
zoo_hc_classes <- stats::cutree(zoo_hc, k = 3)
```
This suggests 2 or 3 clusters, we go with 3.

```{r}
set.seed(42)
zoo_hc_div <- cluster::diana(zoo_gower)
plot(
  zoo_hc_div, which.plots = 2, labels=FALSE, 
  main = "Divisive Clustering on Categorical Data")
rect.hclust(as.hclust(zoo_hc_div), k = 3, border = "red")
zoo_hc_div_classes <- stats::cutree(as.hclust(zoo_hc_div), k = 3)
```

```{r echo=FALSE}
table(true_labels_zoo, zoo_hc_div_classes) |> 
  knitr::kable(
    caption = "Zoo: Predicted clusters HC-DIV k=3"
  )

table(true_labels_zoo, zoo_hc_classes) |> 
  knitr::kable(
    caption = "Zoo: Predicted clusters HC-AGG k=3"
  )
zoo_performance <- zoo_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "HC-AGG",
      variables = "nominal",
      parameters = "k=3",
      rand.seed = "42",
      homo_score = clevr::homogeneity(
        true_labels_zoo, zoo_hc_classes) |> round(2)
      ),
      dplyr::tibble(
      method_name = "HC-DIV",
      variables = "nominal",
      parameters = "k=3",
      rand.seed = "42",
      homo_score = clevr::homogeneity(
        true_labels_zoo, zoo_hc_div_classes) |> round(2)
      )
    )
knitr::kable(
  zoo_performance,
  caption="Zoo: Methods performance")
```


## Zoo: LCM
```{r echo=FALSE}

names_possible_models <- c("Binary_p_E","Binary_p_Ej", "Binary_p_Ek",
                           "Binary_p_Ekj","Binary_p_Ekjh","Binary_pk_E",
                           "Binary_pk_Ej","Binary_pk_Ek","Binary_pk_Ekj",
                           "Binary_pk_Ekjh")

names(names_possible_models) <- names_possible_models
set.seed(42)
zoo_lcm_models <-
  purrr::map(names_possible_models,function(x){
    res <- Rmixmod::mixmodCluster(
      zoo_features,
      nbCluster = 1:12,
      dataType = "qualitative",
      model = Rmixmod::mixmodMultinomialModel(listModels = x),
      criterion = c("ICL")
    )
    res
  })

zoo_lcm_models_lcl <-
  purrr::map(
    names_possible_models,function(model_name){
      
      this_model_results <- zoo_lcm_models[[model_name]]
      
      this_model_nbcluster <- 
        purrr::map_int(1:12,function(x = .x){
          this_model_results@results[[x]]@nbCluster[1]  
        })
      
      this_model_icl <-
        purrr::map_dbl(1:12,function(x = .x){
          this_model_results@results[[x]]@criterionValue[1]
        })*-1
      
      dplyr::tibble(num_clusters = this_model_nbcluster,
                    icl = this_model_icl) |> 
        dplyr::rename(!!model_name := icl)
    }
    
    
  )

zoo_lcm_models_lcl_df <- purrr::reduce(
  zoo_lcm_models_lcl,
  dplyr::left_join,
  by="num_clusters"
)

model_colors <- c(
  "Binary_p_E"="black","Binary_p_Ej"="red","Binary_p_Ek"="green",
  "Binary_p_Ekj"="blue","Binary_p_Ekjh"="purple",
  
  "Binary_pk_E"="black","Binary_pk_Ej"="red","Binary_pk_Ek"="green",
  "Binary_pk_Ekj"="blue","Binary_pk_Ekjh"="purple"
)

model_linetypes <- c(
  "Binary_p_E"="solid","Binary_p_Ej"="solid","Binary_p_Ek"="solid",
  "Binary_p_Ekj"="solid","Binary_p_Ekjh"="solid",
  
  "Binary_pk_E"="dashed","Binary_pk_Ej"="dashed",
  "Binary_pk_Ek"="dashed","Binary_pk_Ekj"="dashed","Binary_pk_Ekjh"="dashed")

plot_base <-
  ggplot(data=zoo_lcm_models_lcl_df, 
         aes(x=num_clusters))

plot_all_lines <-
  purrr::reduce(names_possible_models,function(p,model_name){
    
    p +
      geom_line(aes(y = rlang::eval_tidy(rlang::sym(model_name)),
                    colour = model_name),
                linetype = model_linetypes[model_name]) + 
      geom_point(aes(y =  rlang::eval_tidy(rlang::sym(model_name))),
                 shape = 0)
  },
  .init = plot_base)

plot_final <-
  plot_all_lines +
  scale_color_manual(name = "Models",
                     values = model_colors) +
  scale_x_continuous(breaks=seq(1,12,1)) +
  xlab(label = "Number of clusters") +
  ylab(label = "") +
  labs(
    title = "ICL of 10 Latent Class Models",
    subtitle = "Zoo categorical data"
  ) +
  theme_minimal()   

plot_final

```

```{r echo=FALSE}
zoo_lcm_models_best <- zoo_lcm_models[["Binary_pk_Ej"]]@bestResult 

table(
  true_labels_zoo,
  zoo_lcm_models_best@partition) |> 
  knitr::kable(caption = "Zoo: Predicted clusters LCM Binary_pk_Ej")

zoo_performance <- zoo_performance |> 
  dplyr::bind_rows(
    dplyr::tibble(
      method_name = "LCM",
      variables = "nominal",
      parameters = paste0(
        "k=",
        zoo_lcm_models_best@nbCluster,
        ",",
        zoo_lcm_models_best@model,collapse = ""),
      rand.seed = "42",
      homo_score = clevr::homogeneity(
        true_labels_zoo, zoo_lcm_models_best@partition) |> round(2)
      )
    ) |> 
  dplyr::arrange(dplyr::desc(homo_score))
knitr::kable(
  mushroom_performance,
  caption="Zoo: Methods performance")
```

The LCM outperforms.

# References
